Icicle: write once, run once

### 2016-05-30 2016fhpc/section/02-Source.tex
Copy edits to bounded buffer restriction

\eject
% ---------------------------------------------------------
\section{Elements and Aggregates}

To allow incremental computation all Icicle queries must execute in a single pass over the input stream.
Of course, not all queries \emph{can} be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order different to what the stream provides.
However, as we saw in the introduction, although a particular \emph{algorithm} may be impossible to evaluate in a streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
Here is the unstreamable example from the introduction again:
\begin{code}
  table kvs { key : Date; value : Real }
  query avg = let k = last key
              in  filter (key == k) of mean value;
\end{code}

@@

% ---------------------------------------------------------
\subsection{The Stage Restriction}
To ensure that Icicle queries can be evaluated in a single pass, we use a modal type system inspired by staged computation~\cite{davies2001modal}. We use two modalities, @Element@ and @Aggregate@. Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are only available once the entire stream has been consumed. In the expression @(filter (key == k) of mean value)@, the variable @key@ has type @Element Date@ while @k@ has type @Aggregate Date@. Attempting to compile this query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.

Note that the types of pure values such as constants are automatically promoted to have the required modality, for example, if we have @open == 1@ and @open : Element Int@ then the constant @1@ is automatically promoted to have type @Element Int@ as well.

% ---------------------------------------------------------
\subsection{Finite Streams and Synchronous Data Flow}
As opposed to synchronous data flow languages such as {\sc Lustre}~\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length. Icicle is fundamentally a query language, which queries finite tables of data held in a non-volatile store, but does so in a streaming manner. In contrast, Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like to fly airplanes). In Icicle, the ``last'' element in a stream is the last one that appears in the file on disk. In Lustre, the ``last'' element in a stream is the one that was most recently received. If the ``unstreamable'' version of the query from \S2 was converted into Lustre syntax then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself. This means the filter predicate would always be true, and the query would return the mean of the entire stream. Applying the Icicle type system to our queries imposes the natural stage restriction associated with finite streams, so there are distinct ``during'' (element) and ``after'' (aggregate) stages.

% ---------------------------------------------------------
\subsection{Incremental Update}

Suppose we apply a query to a large table and record the result. Tomorrow morning we receive more data and add it to the table. We would like to recompute the result of this query without needing to process all data from the start of the table. We can do this by remembering the values of all intermediate aggregates that are computed in the query, and updating them as new data arrives. In the @avg@ example from \S2 these aggregates are @k@ and @avgs@.

When exposing impure computations whose value changes across successive runs, these must be bound as aggregates.
For example, we can bind the current time @now@ as an @Aggregate Date@.
This ensures that reused parts of an incremental computation are the same regardless of which day they are executed.

% ---------------------------------------------------------
\subsection{Bounded buffer restriction}
\label{s:IcicleSource:bounded}
Icicle queries are intended to process tables of arbitrary size that may not fit in memory. Due to this, each query must execute without requiring buffer space proportional to the size of the input. For example, here is a simple Haskell expression which cannot be executed without reserving a buffer at least the same size as the input:
\begin{code}
    unbounded (xs :: Stream Int)
     = zip (filter (> 0) xs) (filter (< 0) xs)
\end{code}

The expression takes the input stream @xs@ and selects the elements greater than zero, and pairs them with elements that are less than zero. This expression requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then all the positive values must be retained before any output can be produced.

The language constructs of Icicle are designed so that queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment. In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly. Constructs such as @filter@ and @fold@ do not take the name of the stream as argument, but instead operate on the stream defined in the context. In all the example Icicle queries so far, the name of the table / stream is not mentioned in the body of the query, and serves as a comment only. Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query. In the expression @filter p of mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.

Finally, note that our major caveat is that the @group@ combinator we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream.
For our applications the keys are commonly company names, customer names, and days of the year.
Grouping by types with an unbounded number of members, such as @Real@ or @String@ can be undesirable, and we wish to outlaw this in a future version of our production compiler.

>>>>

\eject
% ---------------------------------------------------------
\section{Elements and Aggregates}
To allow incremental computation all Icicle queries must execute in a single pass over the input stream.
Of course, not all queries \emph{can} be executed in a single pass: the key examples are queries that require random access indexing, or otherwise need to access data in an order different to what the stream provides.
However, as we saw in the introduction, although a particular \emph{algorithm} may be impossible to evaluate in a streaming fashion, the desired \emph{value} may well be computable, if only we had a different algorithm.
Here is the unstreamable example from the introduction again:
\begin{code}
  table kvs { key : Date; value : Real }
  query avg = let k = last key
              in  filter (key == k) of mean value;
\end{code}

@@

% ---------------------------------------------------------
\subsection{The Stage Restriction}
To ensure that Icicle queries can be evaluated in a single pass, we use a modal type system inspired by staged computation~\cite{davies2001modal}. We use two modalities, @Element@ and @Aggregate@. Values of type @Element@~$\tau$ are taken from input stream on a per-element basis, whereas values of type @Aggregate@~$\tau$ are available only once the entire stream has been consumed. In the expression @(filter (key == k) of mean value)@, the variable @key@ has type @Element Date@ while @k@ has type @Aggregate Date@. Attempting to compile the unstreamable query in Icicle will produce a type error complaining that elements cannot be compared with aggregates.

Note that the types of pure values such as constants are automatically promoted to the required modality, for example, if we have @open == 1@ and @open : Element Int@ then the constant @1@ is automatically promoted to have type @Element Int@ as well.

% ---------------------------------------------------------
\subsection{Finite Streams and Synchronous Data Flow}
In contrast to synchronous data flow languages such as {\sc Lustre}~\cite{halbwachs1991synchronous}, all streams processed by Icicle are conceptually finite in length. Icicle is fundamentally a query language, which queries finite tables of data held in a non-volatile store, but does so in a streaming manner. Lustre operates on conceptually infinite streams, such as those found in real-time control systems (like to fly airplanes). In Icicle, the ``last'' element in a stream is the last one that appears in the table on disk. In Lustre, the ``last'' element in a stream is the one that was most recently received. If the unstreamable query from \S2 was converted to Lustre syntax then it would execute, but the filter predicate would compare the last key with the most recent key from the stream, which is the key itself. The filter predicate would always be true, and the query would return the mean of the entire stream. Applying the Icicle type system to our queries imposes the natural stage restriction associated with finite streams, so there are distinct ``during'' (element) and ``after'' (aggregate) stages.

% ---------------------------------------------------------
\subsection{Incremental Update}
Suppose we query a large table and record the result. Tomorrow morning we receive more data and add it to the table. We would like to update the result without needing to process all data from the start of the table. We can do this by remembering the values of all intermediate aggregates that were computed in the query, and updating them as new data arrives. In the @avg@ example from \S2 these aggregates are @k@ and @avgs@.

We can also provide impure contextual information to the query such as the current date, by assigning it an aggregate type. As element-wise computations cannot depend on aggregate computations, we ensure that reused parts of an incremental computation are the same regardless of which day they are executed.

% ---------------------------------------------------------
\subsection{Bounded buffer restriction}
\label{s:IcicleSource:bounded}
Icicle queries process tables of arbitrary size that may not fit in memory. Due to this, each query must execute without requiring buffer space proportional to the size of the input. As a counter example, here is a simple function which cannot be applied without reserving a buffer of the same size as the input:
\begin{code}
    unbounded (xs : Stream Int)
     = zip (filter (> 0) xs) (filter (< 0) xs)
\end{code}

This function takes an input stream @xs@, and pairs the elements that are greater than zero with those that are less than zero.This computation requires an unbounded buffer because if the stream contains $n$ positive values followed by $n$ negative values, then all the positive values must be retained before any output can be produced.

Icicle is designed so that queries that would require unbounded buffering cannot be written, with one major caveat that we will discuss in a moment. In Icicle, the stream being processed (such as @xs@ above) is implicit in the defined query, and is never named explicitly. Constructs such as @filter@ and @fold@ do not take the name of the stream as an argument, but instead operate on the stream defined in the context. In the example Icicle queries presented so far, the name of the table (stream) is not mentioned in the body of the query, and serves as a comment only. Icicle language constructs describe \emph{how elements from the stream should be aggregated}, but the order in which those elements are aggregated is implicit, rather than being definable by the body of the query. In the expression @filter p of mean value@, the term @mean value@ is applied to stream values which satisfy the predicate @p@, but the values to consider are supplied by the context.

Finally, note that our major caveat is that the @group@ construct we used in \S2 uses space proportional to the number of distinct \emph{keys} in the input stream.
For our applications the keys are commonly company names, customer names, and days of the year. Our production system knows that these types are bounded in size and that maps from keys to values will fit easily in memory. Attempting to group by values of a type with an unbounded number of members, such as a @Real@ or @String@ results in a compile-time warning.

% Grouping by types with an unbounded number of members, such as @Real@ or @String@ can be undesirable, and we wish to outlaw this in a future version of our production compiler.

% BEN: "Can be undesirable" is too imprecise. What will happen is that an entry will be added to the map for every key, and if there are more keys than will fit in memory then the query will run out of memory. Surely adding the mentioned warning to the compiler would not be difficult? If it's easy then saying it's done sounds much better than "we intend to". We need to have a solid story about this point. One of the key contributions of Icicle is that it does not require ``unbounded buffering'', but if you're grouping by arbitrary values from the input stream then it clearly does.
